---
title: "An analysis of Adelie bodymass and flipper length"
output: html_document
date: "2023-11-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction

In this code, I explore the palmerpenguins data, looking at the relationship between body mass and flipper length in Adelie penguins. This code specifically tests whether body mass can be used to predict flipper length in this species by performing a linear regression.

```{r packages, include=FALSE, warning=FALSE}

library(palmerpenguins)
library(tidyverse)

sink(file = "package-versions.txt")
sessionInfo()
sink()

```

```{r functions, echo=FALSE}

#Loading in functions
source("functions/Adelie_analysis_functions.r")

```

```{r Creating Dataset, echo=FALSE}

#Saving the original dataset
write.csv(penguins, "data/penguins.csv")

#Creating a new dataset with just the data from Adelie penguins, and removing rows with 'NA's
adelies <- (penguins) %>%
  remove_NA() %>%
  filter_by_adelie()

#Saving the new dataset
write.csv(adelies, "data/adelies.csv")

adelies

```

```{r Data Exploration, echo=FALSE, warning=FALSE}

#Plotting body mass against flipper length for Adelies
exploring_plot <- ggplot(adelies, aes(body_mass_g, flipper_length_mm))+
  geom_point()+
  labs(x = 'body mass (g)', y = 'flipper length (mm)', title = 'Adelie body mass against flipper length')+
  theme_bw()

#Saving the graph
ggsave(filename = "figures/data_exploration.png",
       plot = exploring_plot,
       height = 8 ,
       width = 15 ,
       units = "cm")

exploring_plot

```

### Hypotheses

The above graph shows the data for Adelie penguins in the palmerpenguins package. My null hypothesis is that flipper length in these penguins is not predicted by body mass, and so a linear regression model will not be statistically significant. My alternative hypothesis is that body mass does predict flipper length in this species, and so the regression model will be signficant.

### Statistical Methods

To test whether body mass does predict flipper length in Adelie penguins, I have carried out a linear regression on this data. I have ensured that the assumptions of a linear regression are met, plotting a qqplot to check for normally distributed residuals. I have then used an ANOVA to test overall statistical significance of the regression line by comparing it to a null model.

```{r Statistics}

#Creating a linear model for body mass predicting flipper length in Adelies
model1 <- lm(flipper_length_mm ~ body_mass_g, adelies)

#Getting the equation of the linear model, and R^2 value
summary(model1)

#Creating a QQ plot to look at whether residuals are normally distributed
create_QQplot(model1)
#The points on the QQplot are fairly close to the line throughout and so appear normally distributed, so no transformations of the data are needed

#Doing an ANOVA to look at statistical significance of the model
anova(model1)

```

### Results & Discussion

Firstly the QQplot produced showed that the residuals of this model were normally distributed, and the data also follows the other assumptions of a linear regression and so no transformation was needed. The equation of the regression line of body mass against flipper length is y = 0.007x + 165.2 which shows a positive relationship between the two variables. As body mass increases by 1g, flipper length increases by 0.007mm. We can see that both the gradient and the intercept are significantly different to 0, both with a p value of less than 0.05.

I then did an ANOVA to test the statistical significance of the regression between body mass and flipper length, which resulted in a p value of less than 0.001, showing that we can reject the null hypothesis and accept the alternative hypothesis, that body mass predicts flipper length. The R2 value of the model was 0.214, showing that body mass explains 21% of variation in flipper length. This is a moderate amount, but suggests there may also be other factors which explain variation in flipper length and so by itself, body mass isn't a perfect predictor of flipper length.

```{r Plotting Results, warning=FALSE}

result_plot <- ggplot(adelies, aes(body_mass_g, flipper_length_mm))+
  geom_point(alpha = 0.2)+
  labs(x = 'body mass (g)', y = 'flipper length (mm)', title = 'Adelie body mass against flipper length')+
  theme_bw()+
  geom_smooth(method = 'lm', fill = '#81DBFF')+
annotate(geom = "label", x = 4400, y = 188, label = "y = 0.007x + 165.2, p = < 0.001")

ggsave(filename = "figures/linear_regression_result.png",
       plot = result_plot ,
       height = 8 ,
       width = 15 ,
       units = "cm")

result_plot

```

### Conclusion

The linear regression carried out on body mass data and flipper length data collected from population of Adelie penguins has shown we can use body mass of the penguins to predict their flipper length. The regression and subsequent ANOVA showed a statistically significant positive relationship between the two variables, meaning heavier penguins having longer flippers, however the R2 value of this regression showed that body mass is only a moderate predictor of flipper length, and there are likely other variables which explain more of the flipper length variation seen.
